Sampling Strategies in Decisions from Experience

### 2022-02-16 manuscript.Rmd
Minor Reviews

Now, more formally,^[Paragraphs with this beginning usually formalize the main ideas of the preceding paragraphs or elaborate on them by using mathematical expressions and concepts. Although important, the gist of this paper should be captured despite skipping these paragraphs.] let us treat the choice alternatives as elements $c$ of the set $C$.
To each $c \in C$ is then assigned some information from the inputs and this information may be used by the mind to learn about the properties of the respective alternative.
Accordingly, decision making can be considered an information-processing problem which starts from the set $C$, containing all choice alternatives, and stops at a proper subset $C' \subset C$ with the inclusion map
$$
\begin{aligned}
  \iota : C' &\to C \\
  c &\mapsto c  \quad \forall c \in C'
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, p. 15], containing only the chosen alternatives.

@@

This paper uses the choice pattern in DfE as an entry point for an inquiry into the sampling process as a consequential part of the information processing upfront choices in DfE.
More specifically, it will be discussed and demonstrated in a simulation how changes to the strategy of sequentially generating outcomes and integrating them into samples (hereafter: sampling strategies) can produce distinct choice patterns in DfE---among them, the as-if underweighting of rare outcomes [see @hillsInformationSearchDecisions2010, for the original formulation of this idea].

The rest of the paper is structured as follows.
In the first brief section, a formal definition of DfE is provided.
In the subsequent section, cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] is reviewed as a data model that can be used to describe choices in DfE in terms of deviations from the principle of mathematical expectation.
<!-- Outline of the section -->
In the section that follows, sampling strategies are treated in quite some detail.
<!-- Outline of the section -->
Thereafter, a simulation is presented in which the theoretical claims that are made in the preceding section are implemented in a computational model to simulate choice data in DfE.
The predictions of the model are summarized by fitting the choice data in CPT.
The paper closes with a summary and a conclusion.
Note that this paper, including the simulations and data analyses, is reproducible---materials and instructions can be found on the GitHub repository.

@@

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, Chapter 1, for a discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered the possible outcomes of a choice alternative---hereafter $x_i$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p_X(x_i)$.
Roughly speaking, then, all choice alternatives that can be fully described by their outcome-probability pairs $(x_i, p_X(x_i))$ are prospects (or: gambles), which are exclusively considered below. 
<!--For illustration, take the choice alternatives in the following example with monetary outcomes:

@@

Now, more formally, let the axiomatic definition of a prospect be that of a discrete random variable's distribution.
Let therefore the tuple $(\Omega, \mathcal{F}, P)$ be a probability space. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ is the sample space containing a finite set of elements
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega) = \mathcal{F}$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \to [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.<!--^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \to [0,1]$, which is used to construct the probability measures $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]-->
The random variable $X$ is the measurable function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{4}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ being a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the probability measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{5}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(X^{-1} \ A') = P(\{\omega: X(\omega) = x \in A'\})
\tag{6}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21; see also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1, for a comprehensive introduction to random variables].
That is, the set consisting of a single outcome $x$ can be treated as a subset $\{x\} \in \mathcal{F'}$ and we therefore obtain $p_X(x) = P(X(\omega) = x)$.
Note that the probability measure in (5) and (6), respectively, is a discrete random variable's distribution and therefore a prospect.
To summarize and simplify, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
C = \{P_{X_1}, ... , P_{X_k}\}
\tag{7}
$$
with each element $c \in C$ being a distribution $P_X$ of a random variable defined on a different probability space.
Accordingly, each $c$ can itself be considered a set of outcome-probability pairs $(x_i, p_X(x_i)) \in \{(x_1, p_X(x_1)), ..., (x_n, p_X(x_n))\}$.

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes-probability pairs.
In such cases, people would make *decisions from description* (DfD), since the information from which the mind learns about the prospects are complete descriptions thereof.
Yet, rather often, people make *decisions from experience* [DfE, @hertwigDecisionsExperienceEffect2004], where the mind can learn about the prospects only from a record of past experiences with them.
That is, people may have experienced the possible outcomes of a prospect when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular prospect is a rough description of a sample.
The mind may then infer from the relative frequency with which an outcome occurred in such a sample of past decisions its probability of occurrence following a future decision.

Now, more formally, let the sample of a prospect be a set of realizations of the respective random variable $X$ according to its distribution $P_X$.^[Note that this paper considers the counterfactual case where samples are random, although it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. To account for non-random sampling, e.g., in the form of sampling biases [e.g., @fiedlerBewareSamplesCognitiveecological2000] or dynamic information structures [e.g., @cohenEffectPerceivedPatterns2021; @plonskyRelianceSmallSamples2015], the i.i.d-assumption would need to be dropped, which is out of the scope of this paper.]
So, for the choice between $1, ..., k$ prospects, the inputs to DfE are generated by a set $S$ of stochastic processes
$$
S = \{X_{1t}, ... , X_{kt}\}
\tag{8}
$$
with $t \in \{1, ..., N\}$ denoting a sequence of independent and identically distributed (i.i.d.) random variables and $N$ denoting the size of the respective sequence.
The sets of realizations that result from these stochastic processes may then be used to make an uncertain inference about the respective sets of outcome-probability pairs in (7).

Of course, the accuracy of such inferences depends, ceteris paribus, on the size of the samples, since from the laws of large numbers it follows that the relative frequency with which an outcome occurs in a sample converges to the outcome's objective probability only as the size of a sample increases [see @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 5].
However, if the choices that people actually make between prospects are described, robust differences can be found between the choices in DfD and DfE, and the size of samples in DfE provides just one and only a partial explanation for this *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009; see also @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-anayltic review]. 
The following section briefly addresses how choices between prospects in general and the description-experience gap in particular can be described using cumulative prospect theory's [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] *weighting function*.^[This paper also considers CPT's value function, but emphasis is put on the weighting function.]
<!-- The two sections thereafter ... -->

# The Description of Choices in CPT and As-If Underweighting in DfE

@@

To *describe* how people's actual choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing the value

@@

Below, CPT will therefore only be used as a model to describe choices between prospects and it will be discussed---and demonstrated in a simulation---how distinct strategies of generating and integrating samples in DfE can contribute to weighting patterns that are commonly inferred when modeling respective empirical choice data.

@@

(ref:weighting) Probability Weighting Function of @goldsteinExpressionTheoryPreference1987 for Different Values of the Curvature Parameter $\gamma$ and the Elevation Parameter $\delta$.

@@

ggplot(wf) +
  geom_line(aes(prob, w, group = gamma, color = gamma)) +
  geom_segment(x = 0, xend = 1, y =  0, yend = 1, linetype = "dashed", size = .2, color = "gray") +
  scale_color_scico(palette = "lisbon") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x = "p",
       y = "w(p)", 
       color = expression(gamma)) + 
  facet_wrap(. ~ delta, labeller = label_bquote(cols = delta: .(delta))) + 
  theme_minimal()
```

@@

cpt_w %>% 
  group_by(boundary, a, s) %>% 
  ggplot(aes(x = ep, y = w, color = s)) +
  geom_path() + 
  scale_color_scico(palette = "grayC") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x ="Experienced Probability",
       y = "Probability Weighting", 
       color="Switching Probability") + 
  theme_minimal()

cpt_w %>% 
  group_by(boundary, a, s) %>% 
  ggplot(aes(x = ep, y = w, color = s)) +
  geom_path() + 
  scale_color_scico(palette = "buda") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  facet_wrap(~a) +
  labs(x ="Experienced Probability",
       y = "Probability Weighting", 
       color="Switching Probability") + 
  theme_minimal()
```

>>>>

Now, more formally,^[Paragraphs with this beginning usually formalize the ideas of the preceding paragraphs or elaborate on them by using mathematical expressions and concepts. Although I consider these paragraphs important, the major propositions of this paper should be captured despite skipping them.] let us treat the choice alternatives as elements $c$ of the set $C$.
To each $c \in C$ is then assigned some information from the inputs and this information may be used by the mind to learn about the properties of the respective alternative.
Accordingly, decision making can be considered an information-processing problem which starts from the set $C$, containing all choice alternatives, and stops at a proper subset $C' \subset C$ with the inclusion map
$$
\begin{aligned}
  \iota : C' &\to C \\
  c &\mapsto c  \quad \forall c \in C'
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, p. 15], containing only the chosen alternatives.

@@

This paper uses the as-if underweighting of rare outcomes as a starting point for an inquiry into the sampling process upfront choices in DfE.
Essentially, it discusses and demonstrates in a simulation how different strategies of sequentially generating outcomes and integrating them into samples (hereafter: sampling strategies) can produce distinct choice patterns in DfE---among them, the as-if underweighting of rare outcomes [see @hillsInformationSearchDecisions2010, for the original formulation of this claim].
<!-- It thereby proposes that ... -->

The rest of the paper is therefore structured as follows.
In the first brief section, a formal definition of DfE is provided.
In the subsequent section, cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] is reviewed as a data model that can be used to describe choices in DfE in terms of deviations from the principle of mathematical expectation.
<!-- Outline of the section -->
In the section that follows, sampling strategies are treated in quite some detail.
<!-- Outline of the section -->
Thereafter, a simulation is presented in which the theoretical claims that are made in the preceding section are implemented in a computational model to simulate choice data in DfE.
The predictions of the model are summarized by fitting the choice data in CPT.
The paper closes with a summary and a conclusion.
Note that this paper, including the simulations and data analyses, is reproducible---materials and instructions can be found on the GitHub repository.

@@

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, Chapter 1, for a discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered the possible outcomes of a choice alternative (hereafter: $x_i$) and the probabilities with which these outcomes occur following the choice of the alternative (hereafter: $p_X(x_i)$).
Roughly speaking, then, all choice alternatives that can be fully described by their outcome-probability pairs $(x_i, p_X(x_i))$ are prospects (or: gambles), which are exclusively considered below. 
<!--For illustration, take the choice alternatives in the following example with monetary outcomes:

@@

Now, more formally, let the axiomatic definition of a prospect be that of a discrete random variable's distribution.
Let therefore the tuple $(\Omega, \mathcal{F}, P)$ be a probability space. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ is the sample space containing a finite set of elements
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega) = \mathcal{F}$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \to [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.<!--^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \to [0,1]$, which is used to construct the probability measures $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]-->
The random variable $X$ is the measurable function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{4}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ being a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the probability measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{5}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(X^{-1} \ A') = P(\{\omega: X(\omega) = x \in A'\})
\tag{6}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21-22].^[See also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1, for a comprehensive yet accessible introduction to random variables.]
That is, the set consisting of a single outcome $x$ can be treated as a subset $\{x\} \in \mathcal{F'}$ and we therefore obtain $p_X(x) = P(X(\omega) = x)$.
Note that the probability measure in (5) and (6), respectively, is a discrete random variable's distribution and therefore a prospect.
To summarize and simplify, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
C = \{P_{X_1}, ... , P_{X_k}\}
\tag{7}
$$
with each element $c \in C$ being a distribution $P_X$ of a random variable defined on a different probability space.
Accordingly, each $c$ can itself be considered a set of outcome-probability pairs $(x_i, p_X(x_i)) \in \{(x_1, p_X(x_1)), ..., (x_n, p_X(x_n))\}$.

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes-probability pairs.
In such cases, people would make decisions from description (DfD), since the information from which the mind learns about the prospects are complete descriptions thereof.
Yet, rather often, people make decisions from experience (DfE), where the mind can learn about the prospects only from a record of past experiences with them.
That is, people may have experienced the possible outcomes of a prospect when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular prospect is a rough description of a sample.
The mind may then infer from the relative frequency with which an outcome occurred in such a sample of past decisions its probability of occurrence following a future decision.

Now, more formally, let the sample of a prospect be a set of realizations of the respective random variable $X$ according to its distribution $P_X$.^[Note that this paper considers the counterfactual case where samples are random, although it has been emphasized that in the natural world samples are "virtually never random" [@fiedlerBewareSamplesCognitiveecological2000, p. 660]. To account for non-random sampling, e.g., in the form of sampling biases [e.g., @fiedlerBewareSamplesCognitiveecological2000] or dynamic information structures [e.g., @cohenEffectPerceivedPatterns2021; @plonskyRelianceSmallSamples2015], the i.i.d-assumption would need to be dropped, which is out of the scope of this paper.]
So, for the choice between $1, ..., k$ prospects, the inputs to DfE are generated by a set $S$ of stochastic processes
$$
S = \{X_{1t}, ... , X_{kt}\}
\tag{8}
$$
with $t \in \{1, ..., N\}$ denoting a sequence of independent and identically distributed (i.i.d.) random variables and $N$ denoting the length of the respective sequence.
The sets of realizations that result from these stochastic processes may then be used to make an uncertain inference about the respective sets of outcome-probability pairs in (7).

Of course, the accuracy of such inferences depends, ceteris paribus, on the size of the samples, since from the laws of large numbers [see, e.g., @kolmogorovFoundationsTheoryProbability1950] it follows that the relative frequency with which an outcome occurs in a sample converges to the outcome's objective probability only as the size of a sample increases.
Yet, if the choices that people actually make between prospects in DfD and in DfE are described, the differences that were already outlined in the introduction are found, and the size of samples in DfE provides just one and only a partial explanation for this so-called *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009; see also @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-analytic review].
This paper will address this issue<!-- review-->.
Before that, however, the following section reviews how choices between prospects can be described using cumulative prospect theory's (CPT) *weighting function*.^[This paper also considers CPT's value function, but emphasis is put on the weighting function.]

# The Description of Choices in CPT and the As-If Underweighting in DfE

@@

To *describe* how people's actual choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of CPT [@tverskyAdvancesProspectTheory1992] and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing the value

@@

Below, CPT will therefore only be used as a data model to describe choices between prospects and it will be discussed---and demonstrated in a simulation---how distinct strategies of generating and integrating samples in DfE can contribute to weighting patterns that are commonly inferred when modeling respective empirical choice data.

@@

(ref:weighting) Possible Shapes of Goldstein and Einhorn's -@goldsteinExpressionTheoryPreference1987 Weighting Function

@@

ggplot(wf) +
  geom_line(aes(prob, w, group = gamma, color = gamma)) +
  geom_segment(x = 0, xend = 1, y =  0, yend = 1, linetype = "dashed", size = .2, color = "gray") +
  scale_color_scico(palette = "cork") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x = "p",
       y = "w(p)", 
       color = expression(gamma)) + 
  facet_wrap(. ~ delta, labeller = label_bquote(cols = delta: .(delta))) + 
  theme_minimal()
```

@@

cpt_w %>% 
  group_by(boundary, a, s) %>% 
  ggplot(aes(x = ep, y = w, color = s)) +
  geom_path() + 
  scale_color_scico(palette = "tokyo") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  labs(x ="Experienced Probability",
       y = "Probability Weighting", 
       color="Switching Probability") + 
  theme_minimal()

cpt_w %>% 
  group_by(boundary, a, s) %>% 
  ggplot(aes(x = ep, y = w, color = s)) +
  geom_path(size = 1) + 
  scale_color_scico(palette = "tokyo") + 
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  scale_y_continuous(breaks = seq(0, 1, .2)) + 
  facet_grid(boundary~a) +
  labs(x ="Experienced Probability",
       y = "Probability Weighting", 
       color="Switching Probability") + 
  theme_dark()
```
